<style>
div.purple pre { background-color:purple;}
div.purple pre.r { background-color:purple;}
</style>

# R at GAD {#r_at_GAD}

The following are the GAD [sensible defaults](#sensible_defaults) for R:

## R Version & IDE
The dominant IDE for R is Rstudio, which comes packaged with R. 
For a new project you should use the latest version of Rstudio available from the software portal.

## General
Default to packages from the [Tidyverse](http://tidyverse.org/).These have been carefully designed to work together effectively as part of a modern data analysis workflow. More info can be found here: [R for Data Science by Hadley Wickham](http://r4ds.had.co.nz). 

For example:

   * Prefer tibbles to data.frames
   * Use ggplot2 rather than base graphics
   * Use the pipe `%>%` rather than nesting function calls. (...but not always e.g. see [here](https://twitter.com/hadleywickham/status/603883121197514752)).
   * Prefer `purrr` to the `apply` family of functions. See [here](http://r4ds.had.co.nz/iteration.html#the-map-functions)
   

## Packages {#r_default_packages}

Recommended Packages: 

* Data Manupulation and Tidying
    * [dplyr](https://dplyr.tidyverse.org/)
    * [tidyr](https://tidyr.tidyverse.org/)
* Working with messy spreadsheets
    * [tidyxl](https://nacnudus.github.io/tidyxl/)
    * [unpivotr](https://github.com/nacnudus/unpivotr)
* Working with large datasets:
    * [data.table](https://rdatatable.gitlab.io/data.table/)
    * [dtplyr](https://dtplyr.tidyverse.org/)
* Database Interface
    * [dbplyr](https://dbplyr.tidyverse.org/)
* Plotting
    * [ggplot2](https://ggplot2.tidyverse.org/)
    * [plotly](https://plot.ly/r/)
* Mapping
    * [sf](https://github.com/r-spatial/sf/)
    * [leaflet](http://rstudio.github.io/leaflet/)
* Reproducibility
    * [renv](https://rstudio.github.io/renv/)
* Producing gov.uk style output
    * [govstyle](http://ukgovdatascience.github.io/govstyle/index.html) (Government Theme for ggplot2)
    * [govdown](https://ukgovdatascience.github.io/govdown/)
* Machine Learning
    * [tensorflow](https://tensorflow.rstudio.com/)
    * [h2o](https://github.com/h2oai/h2o-3)
  

## Project Workflow {#r_projects}
Always work in a project. See the guide to [Using Projects](https://support.rstudio.com/hc/en-us/articles/200526207-Using-Projects).

Projects functionality is broken in GAD's packaged version of Rstudio - see the fix [here](R/fix_projects.R)

## Packaging Your Code {#r_package}
Packages are the fundamental unit of reproducible R code. 
Therefore, if possible, build an R Package to share and document your code. 

Hadley's book on [R Packages](http://r-pkgs.had.co.nz/) is an effective guide on how to produce a package.

The [usethis](https://usethis.r-lib.org/) package has lots of useful shortcuts for package builders.

## Managing Dependencies {#r_dependencies}
There are two key competing ways of managing dependencies for an R Project:

* [`packrat`](https://rstudio.github.io/packrat/) - current established way to manage R dependencies
* [`renv`](https://rstudio.github.io/renv/) - rapidly maturing, successor to packrat.

See also:

* Rstudio page on [managing environments in R](https://environments.rstudio.com/)
* Rbloggers [post on renv](https://www.r-bloggers.com/reproducible-environments/)
* Rstudio page on combining [renv and docker](https://rstudio.github.io/renv/articles/docker.html)

### Using old versions of packages {#r_checkpoint}
You may come across code which doesn't work because it depends on a different version of a package to the one you have.

Fortunately, Microsoft keep daily snapshots of CRAN and store them on the [Microsoft R Application Network](https://mran.microsoft.com/timemachine).

The `checkpoint` package from Microsoft lets you use these snapshots to install packages as if it were any day since 2017-07-01.

Simply start your script with:
```{r eval=FALSE}
library(checkpoint)

checkpoint(snapshotDate = "2015-01-15",
           checkpointLocation = getwd()) 
```

This will download and fetch all the packages as they existed on the given date and install them to a library on your home drive.

Notes:

* If the code depends on `BH` (a lot of tidyverse code will) then this will take some time!
* By default checkpoint puts packages on your P drive - this will be slow.
    * You can use the `checkpointLocation` argument to tell checkpoint to use the C drive.

## Error Handling {#r_errorhandling}

Base R includes the `try()` and `tryCatch()` functions for handling errors. You can find an example of basic use of these on [r-bloggers](https://www.r-bloggers.com/error-handling-in-r/).

Effective error handling in R requires understanding the _conditions_ system. There is a good chapter on this in Hadley's [Advanced R book](https://adv-r.hadley.nz/conditions.html)

If you are iterating over many inputs, it is recommended that you use the `safely()` family of functions from `purrr` to create versions which return errors within a list for handling at a later stage.

## Unit Testing {#r_tests}
Use the `testthat` package for performing unit tests. 
For details see the ['tests' chapter of R Packages](https://r-pkgs.org/tests.html).


## Assignment {#r_assignment}

### The `<-` and `->` operators

Assignment should always involve assigning a value on the right-hand-side of the `<-` operator to the object on its left hand side. The `->` operator should *never* be used, and the `=` operator should not be used for assignment.





```{r  eval = FALSE, class.source="Rchunksk", class.output="Routsk"}

#good
a <- 1

#bad
1 -> a

#also bad
a = 1
```


An exception is where the `assign()` function or `<<-` are used, in which case 
assignment can be done without `<-` or `->`. 
This should only be done when `<-` cannot be used.
And as with the above, never use `->>` instead of `<<-`  .

Note - if the code below is confusing, section 10.2.4
here may be of use:
https://adv-r.hadley.nz/function-factories.html#stateful-funs


```{r  eval = FALSE, class.source="Rchunksk", class.output="Routsk"}

#only do this when `<-` will not work, e.g. when
#the name of the object being assigned to is stored 
#as a string:
assign('a', 1, envir = globalenv())

#if assigning to the scope of the enclosing environment, 
#use `<<-` 
create_counter <- function() {
  i <- 0
  
  function() {
    i <<- i + 1
    return(i)
  }
}

#but don't use `<<-` when `<-` would work just as well!

#good:
a <- 1

#bad:
a <<- 1

#even worse:
1 ->> a


```


## Commenting {#r_commenting}

Code should always be well commented; the sections below provide
guidance on good commenting practice.

### Position of comments

Comments should appear above related sections of code.

```{r  eval = FALSE, class.source="Rchunksk", class.output="Routsk"}

#Good:

  #assigning value of 1 to object a:
  a <- 1

#Bad:

  a <- 1 #assigning value of 1 to object a


```


### Detail of comments

It is generally preferable to err on the side of including more comments rather than not commenting enough.
Nevertheless, you should be judicious in your use of comments - adding a comment to explain something 
obvious will unnecessarily clutter your code. The most important places to include comments are where
the code is doing something complicated and non-obvious, particularly if it is using functions or syntax that
others may not be familiar with. This being the case, comment usage should sometimes be adjusted based on
the technical proficiency of others working on the project: if you know that the code will be checked by 
someone who has a lot of experience in R, you will rarely need comments that help the reader to understand the 
syntax being employed. Conversely, if it will be checked by a relative novice, it may be helpful to provide 
more details comments about the operations being performed by the code:



```{r eval = FALSE, class.source="Rchunksk", class.output="Routsk"}

#The code chunk below changes the format of a dataframe 
#from a 'wide' format to a 'long' format. The extent 
#of commenting required may depend on the technical
#proficiency of other people who will work with the 
#code. If this is subject to uncertainty, err on the
#side of more thorough explanations

data.frame(
  animal_id = 1:10, 
  weight = rnorm(n = 10, mean = 5, sd = 3), 
  height = rnorm(n = 10, mean = 7, sd = 3) 
  ) %>%
  pivot_longer(
    c(weight, height)
  )

```


### Outdated comments

When making changes to code, it is important to correspondingly update comments.
Otherwise, comments may become misleading:

```{r  eval = FALSE, class.source="Rchunksk", class.output="Routsk"}

#Uh-oh! Somebody has updated the 'convert_units()' 
#function so that it converts from grams to 
#kilograms instead of  grams to tonnes. But they
#didn't update the comment!

  #The function below converts from grams to tonnes:
  convert_units <- function(value) {
    value <- value / 1000
    return(value)
  }


```

### Different commenting contexts

In some cases, such as when commenting non-code sections of Rmarkdown documents
or pre-pending comments to functions within your own packages using roxygen 
or similar, the above conventions will not suffice because different commenting
syntax will be involved. These situations are likely to be rare in GAD's work 
and are thus beyond the scope of this guidance.

## Indentation and spacing {#r_indentation_and_spacing}

When coding in R, indentation and spacing generally has 
absolutely no impact on the way in which code is evaluated.
We should use the freedom that this allows us to space 
and indent code in a manner that maximises readability.
A few good practices to follow are:

1. Opening brackets should be followed by a new line
plus tab indendtation
2. Closing brackets should be preceded by a new line
plus removal of tab indentation
3. We can make exceptions to the above when brackets 
only enclose a small expression
4. The `<-`,  `=` and `<<-` operators should be followed 
and preceded by a space, as should commas
5. Within tidyverse and magrittr, pipe operators 
like `%>%`, `%$%` and `%T>%` should come at the
end of a line

```{r eval = FALSE, class.source="Rchunksk", class.output="Routsk"}

#good:
data <- data.frame(
  animal_id = 1:10, 
  #it's OK to have rnorm() on a single line because the
  #brackets don't enclose much code.
  weight = rnorm(n = 10, mean = 5, sd = 3), 
  height = rnorm(n = 10, mean = 7, sd = 3) 
  ) %>%
  pivot_longer(
    c(weight, height)
  )

#bad:
data<-data.frame(animal_id=1:10,weight = rnorm(n = 10, 
mean = 5, sd = 3),height=rnorm(
  n = 10, mean = 7,sd=3
)) %>% pivot_longer(c(weight, height)
)

```

## Data management {#r_data_management}

If possible, it is better to store data in tidy dataframes rather than, 
for example, in nested lists:


```{r eval = FALSE, class.source="Rchunksk", class.output="Routsk"}

#good:
data <- data.frame(
  gender = c('male', 'male','female','female'), 
  scheme = c('legacy', 'care', 'legacy','care'),
  value = c(1000, 2000, 3000, 4000)
)

#bad:
data <- list(
  male = list(legacy = 1000, care = 2000),
  female = list(legacy = 3000, care = 4000)
)

#It often makes more sense to create a tibble
#rather than a standard dataframe, but this
#can depend on context:

#also good:
data <- dplyr::tibble(
  gender = c('male', 'male','female','female'), 
  scheme = c('legacy', 'care', 'legacy','care'),
  value = c(1000, 2000, 3000, 4000)
)



```

## Naming conventions {#r_naming_conventions}

Within a given piece of code, all objects and names within vectors, lists or 
dataframes should follow camel case or snake case convention, but not both.
Camel case involves writing names with the first chunk of the name uncapitalised
and all subseqeunt chunks capitalised; a 'chunk' in this context means a part
of the name that would be surrounded by spaces in normal writing. Snake case
involves replacing spaces with underscores and removing all capitalisation:

thisObjectNameIsWrittenInCamelCase

this_object_name_is_written_in_snake_case

In addition, we should try to avoid making names too long (unlike the above), 
but this should be balanced against the need to avoid shortening long strings
into incomprehensible abbreviations

```{r eval = FALSE, class.source="Rchunksk", class.output="Routsk"}


#bad - not camel case or snake case:
my_Object_Nm <- 1
AnotherOBJECT <- 2

#bad - mixing camel case and snake case in same code:
my_object_name <- 1
anotherObject <- 2

#bad - confusing abbreviations or arbitrary names make it
#difficult to easily discern what the objects are supposed
#to represent:
m_o_n <- 1
x <- 2

#good - exclusively camel case, 
#abbreviations comprehensible:
myObjName <- 1
anotherObj <- 2

#good - exclusively snake case, 
#abbreviations comprehensible:
my_obj_name <- 1
another_obj <- 2


```

## Folder structure and path specification {#r_folder_structure_and_path_specification}

In general, R code should be contained within folders that have 
a pre-arranged structure from the outset, allowing for the use of
relative file paths. The advantage of this is that the relevant
folder can contain all code, data and other necessary files and 
can be moved around to different parent folders and locations, without
paths needing to be re-written when, for example, code is sent to a client:

```{r,  eval = FALSE, class.source="Rchunksk", class.output="Routsk"}

###Assume we have a folder structure as follows and that
###the code below is found in 'our_code.R'

#project_folder
#   - data_folder
#     - dataset_1.csv
#     - dataset_2.csv
#   - code_folder
#     - our_code.R
#     - more_code.R

#good:
  
  #load packages
  library(readr)

  #read in data using relative paths
  dataset_1 <- read_csv('
      ../data_folder/dataset_1.csv'
      )
  dataset_2 <- read_csv('
      ../data_folder/dataset_2.csv'
      )
  #create user-defined functions
  source('./more_code.R')
  
#bad:
  
  #load packages
  library(readr)

  #read in data using absolute paths
  dataset_1 <- read_csv(
    '<long_and_convoluted_path>/dataset_1.csv'
    )
  dataset_2 <- read_csv(
    '<long_and_convoluted_path>/dataset_2.csv'
    )
  
  #create user-defined functions
  source('<long_and_convoluted_path>/more_code.R') 


```

## Reading in data from csv files {#r_reading_in_data_from_csv_files}

When reading in data from a csv file, it is preferable to use the `read_csv()`
function from the `readr` package, rather than the base function `read.csv()`
This is because the latter is liable to behave differently for different users, 
see: 
https://r4ds.had.co.nz/data-import.html
